---
title: "TOD in Washington, DC"
author: "Jenna Epstein, David Seunglee Park, Leah Shapiro"
date: "9/25/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---
# Set Up
```{r setup, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
	echo = TRUE)
options(scipen=999)
options(tigris_class = "sf")

library(tidyverse)
library(tidycensus)
library(sf)
library(knitr)
library(kableExtra)
library(rmarkdown)
library(tigris)
library(viridis)

```

## Census API Key
```{r, include=TRUE, message=FALSE, warning=FALSE}
# Make sure to replace this with your own API key! This one is Jenna's.
census_api_key("41e1c0d912341017fa6f36a5da061d3b23de335e", overwrite = TRUE)
```

## Load themes and palettes
```{r load_themes, warning=FALSE, message=FALSE}
mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 16,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.text.x = element_text(size = 14))
}

plotTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 16,colour = "black"),
    plot.subtitle = element_text(face="italic"),
    plot.caption = element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey80", size = 0.1),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.background = element_rect(fill = "grey80", color = "white"),
    strip.text = element_text(size=12),
    axis.title = element_text(size=12),
    axis.text = element_text(size=10),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"),
    strip.text.x = element_text(size = 14)
  )
}
qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],0),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]]), digits = 3),
                 c(.01,.2,.4,.6,.8), na.rm=T)
  }
}

q5 <- function(variable) {as.factor(ntile(variable, 5))}

# Load hexadecimal color palette
# Discrete from YlGnBu colorbrewer palette
# This way also the lightest is yellow, won't get confused with NA
palette5.YlGnBu <- c("#ffffcc","#a1dab4","#41b6c4","#2c7fb8","#253494")

# Here's a palette of two colors for TOD and Non-TOD, similar to what we used in lab.
paletteTOD <- c("#fdb863", "#80cdc1")
```

# Introduction

```{r}

```

# Data Wrangling

```{r data_wrangling, message=FALSE, warning=FALSE, include=TRUE, results='hide'}

# Creating a vector of census variables
acs_vars_DC <- c("B02001_001E", # Estimate!!Total population by race
                "B02001_002E", # People describing themselves as "white alone"
                "B02001_003E", # People describing themselves as "black" or "african-american" alone
                "B15001_050E", # Females with bachelors degrees
                "B15001_009E", # Males with bachelors degrees
                "B19013_001E", # Median HH income
                "B25058_001E", # Median rent
                "B06012_002E", # Total poverty
                "B08301_001E", # People who have means of transportation to work
                "B08301_002E", # Total people who commute by car, truck, or van
                "B08301_010E", # Total people who commute by public transportation"
                "B03002_012E", # Estimate Total Hispanic or Latino by race
                "B19326_001E") # Median income in past 12 months (inflation-adjusted)


# ---- Washington, DC - Census Data - 2009 ----
tracts2009 <- 
  get_acs(geography = "tract", 
          variables = acs_vars_DC, 
          year=2009, 
          state=11, 
          geometry=T, 
          output="wide") %>%
  st_transform('ESRI:102685') %>% # using the state plane of Maryland for a projection
  rename(TotalPop = B02001_001E, 
         Whites = B02001_002E,
         Blacks = B02001_003E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         TotalCommute = B08301_001E,
         CarCommute = B08301_002E,
         PubCommute = B08301_010E,
         TotalHispanic = B03002_012E,
         MedInc = B19326_001E) %>%
  dplyr::select(-NAME, -starts_with("B0"), -starts_with("B1"), -starts_with("B2")) %>%
  mutate(pctWhite = (ifelse(TotalPop > 0, Whites / TotalPop,0))*100,
         pctBlack = (ifelse(TotalPop > 0, Blacks / TotalPop,0))*100,
         pctHis = (ifelse(TotalPop >0, TotalHispanic/TotalPop,0))*100,
         pctBlackorHis = (ifelse (TotalPop>0, (Blacks+TotalHispanic)/TotalPop,0)) *100,
         pctBachelors = (ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0)) *100,
         pctPoverty = (ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0))*100,
         pctCarCommute = (ifelse(TotalCommute > 0, CarCommute / TotalCommute,0))*100,
         pctPubCommute = (ifelse(TotalCommute > 0, PubCommute / TotalCommute,0))*100,
         year = "2009") %>%
  dplyr::select(-Whites, -Blacks, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -CarCommute, -PubCommute, -TotalCommute, -TotalHispanic)

# ---- Washington, DC - Census Data - 2017 ----
tracts2017 <- 
  get_acs(geography = "tract", 
          variables = acs_vars_DC, 
          year=2017, 
          state=11, 
          geometry=T, 
          output="wide") %>%
  st_transform('ESRI:102685') %>%
  rename(TotalPop = B02001_001E, 
         Whites = B02001_002E,
         Blacks = B02001_003E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         TotalCommute = B08301_001E,
         CarCommute = B08301_002E,
         PubCommute = B08301_010E,
         TotalHispanic = B03002_012E,
         MedInc = B19326_001E) %>%
  dplyr::select(-NAME, -starts_with("B0"), -starts_with("B1"), -starts_with("B2")) %>%
  mutate(pctWhite = (ifelse(TotalPop > 0, Whites / TotalPop,0))*100,
         pctBlack = (ifelse(TotalPop > 0, Blacks / TotalPop,0))*100,
         pctHis = (ifelse(TotalPop >0, TotalHispanic/TotalPop,0))*100,
         pctBlackorHis = (ifelse (TotalPop>0, (Blacks+TotalHispanic)/TotalPop,0)) *100,
         pctBachelors = (ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0)) *100,
         pctPoverty = (ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0))*100,
         pctCarCommute = (ifelse(TotalCommute > 0, CarCommute / TotalCommute,0))*100,
         pctPubCommute = (ifelse(TotalCommute > 0, PubCommute / TotalCommute,0))*100,
         year = "2017") %>%
  dplyr::select(-Whites, -Blacks, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -CarCommute, -PubCommute, -TotalCommute, -TotalHispanic)


# ---- Combining 2009 and 2017 data ----
allTracts <- rbind(tracts2009,tracts2017)

# ---- Wrangling Transit Open Data (WMATA)----
# Bringing in data for WMATA stops (stations) 
wmataStops <- 
  rbind(
    st_read("https://opendata.arcgis.com/datasets/54018b7f06b943f2af278bbe415df1de_52.geojson") %>% 
      select(NAME, LINE)) %>%
  st_transform(st_crs(tracts2009))

wmataStops <- wmataStops %>%
  mutate(LINE = strsplit(as.character(LINE), ",")) %>%
  unnest(LINE) %>%
  mutate(across(where(is.character), str_trim)) %>%
  mutate(LINE = replace(LINE,LINE == "slvr", "silver"))%>%
  mutate(LINE = replace(LINE,LINE == "yllw", "yellow"))%>%
  mutate(LINE = replace(LINE,LINE == "grn", "green"))

# Visualizing wmata stops. Note that overlapping lines have overlapping points
ggplot() + 
  geom_sf(data=st_union(tracts2009)) +
  geom_sf(data=wmataStops, 
          aes(colour = LINE),
          show.legend = "point", size= 2, alpha = 0.5) +
  scale_colour_manual(values = c("red" = "red", "orange" = "orange", "yellow" = "yellow", "green" = "green", "blue" = "blue", "silver" = "gray")) +
  labs(title="WMATA Stops", 
       subtitle="Washington, DC", 
       caption="Source: opendata.dc.gov") +
  mapTheme()

# Bringing in data for WMATA lines ----
wmataLines <- 
  rbind(
    st_read("https://opendata.arcgis.com/datasets/a29b9dbb2f00459db2b0c3c56faca297_106.geojson") %>%
      select(NAME)) %>%
    st_transform(st_crs(tracts2009))


# --- Relating WMATA Stops and Tracts ----

# Creating buffers (in feet - note the CRS) around WMATA stops
# Both a buffer for each stop, and a union of the buffers, and then bind these objects together

wmataBuffers <- 
  rbind(
    st_buffer(wmataStops, 2640) %>% # projection is in feet
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
    st_union(st_buffer(wmataStops, 2640)) %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))

# ---- Small multiple facet_wrap plot showing both buffers ----

ggplot() +
  geom_sf(data=wmataBuffers) +
  geom_sf(data=wmataStops, show.legend = "point") +
  facet_wrap(~Legend) +
  mapTheme()

# ---- Spatial operations ----

# sf object with ONLY the unioned buffer
buffer <- filter(wmataBuffers, Legend=="Unioned Buffer")

# Showing that we examine the three different spatial selection types 
# to further illustrate why the select by centroids method is best
# based on tracts2009

# Clip
clip <- 
  st_intersection(buffer, tracts2009) %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Clip")

# Spatial selection
selection <- 
  tracts2009[buffer,] %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Selection")

# Centroid-in-polygon join to see which tracts have their centroid in the buffer
selectCentroids <-
  st_centroid(tracts2009)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(tracts2009, GEOID)) %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")

# Visualizing using Total Population as the fill
allTracts.group.TotalPop <- 
  rbind(clip, selection, selectCentroids)
    
SpatialSelectionTypes2009tracts <-
      ggplot(allTracts.group.TotalPop)+
      geom_sf(data = st_union(tracts2009))+
      geom_sf(aes(fill = TotalPop)) +
      labs(title = "Total Population within 1/2 mi of a WMATA station, 2009", subtitle = "Three types of spatial selection") +
      facet_wrap(~Selection_Type) +
      mapTheme() + 
      theme(plot.title = element_text(size=20))
SpatialSelectionTypes2009tracts

# Centroid joins as above, then "disjoin" to get the ones that *don't* join
# Then, add them all together
# Contains a correct 2009-2017 inflation calculation
allTracts.group <- 
  rbind(
    st_centroid(allTracts)[buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2009", MedRent * 1.14, MedRent)) 

# Visualizing Time/Space Groups
TimeSpaceGrps <-
  ggplot(allTracts.group)+
  geom_sf(data = st_union(tracts2009))+
  geom_sf(aes(fill = TOD)) +
  labs(title = "Time/Space Groups") +
  facet_wrap(~year)+
  scale_fill_manual(values = paletteTOD)+
  mapTheme() + 
  theme(plot.title = element_text(size=22))
TimeSpaceGrps

```

# Four small-multiple (2009 & 2017) visualizations comparing four selected Census variables across time and space (TOD vs. non-TOD)
```{r}

```

## Visualizing Median Rent
```{r task2_MedianRent, warning=FALSE, message=FALSE}
## NEED TO DO: add the wmata line to the legend

mapsMedRent <-
  ggplot(allTracts.group)+
  geom_sf(data = st_union(tracts2009))+
  geom_sf(aes(fill = q5(MedRent.inf)), color = NA, alpha = 0.75) +
  geom_sf(data = buffer, fill = "transparent", color = "red", size = 1.25)+
  geom_sf(data = wmataLines, color = "black", size = 1)+
  scale_fill_manual(values = palette5.YlGnBu,
                    labels = qBr(allTracts.group, "MedRent.inf"),
                    name = "Median Rent ($)\n(Quintile Breaks)") +
  labs(title = "Median Rent by Census Tract, 2009-2017", subtitle = "Real Dollars; Red border denotes areas close to WMATA stations") +
  facet_wrap(~year)+
  mapTheme() + 
  theme(plot.title = element_text(size=22))
mapsMedRent
```

## Visualizing Percent of Total Population who Identify as "White Alone"
```{r task2_PercentWhite, warning=FALSE, message=FALSE}
## NEED TO DO: add the wmata line to the legend

mapsPctWhite <-
  ggplot(allTracts.group)+
  geom_sf(data = st_union(tracts2009))+
  geom_sf(aes(fill = q5(pctWhite)), color = NA, alpha = 0.75) +
  geom_sf(data = buffer, fill = "transparent",color = "red", size = 1.25)+
  geom_sf(data = wmataLines, color = "black", size = 1)+
  scale_fill_manual(values = palette5.YlGnBu,
                    labels = qBr(allTracts.group, "pctWhite"),
                    name = "Percent White\n(Quintile Breaks)") +
  labs(title = "% of White population 2009-2017", subtitle = "Red border denotes areas close to WMATA stations") +
  facet_wrap(~year)+
  mapTheme() + 
  theme(plot.title = element_text(size=22))
mapsPctWhite
```

## Visualizing [VARIABLE 3]
```{r}

```

## Visualizing [VARIABLE 4]
```{r}

```

# Table: Comparing Selected Census Variables across Time and Space (TOD vs. non-TOD)
```{r comparison_table}
allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>%
  group_by(year, TOD) %>%
  summarize(Rent = mean(MedRent, na.rm = T),
            Population = mean(TotalPop, na.rm = T),
            Percent_White = mean(pctWhite, na.rm = T),
            Percent_Black = mean(pctBlack, na.rm = T),
            Percent_HispanicLatino = mean(pctHis, na.rm=T),
            Percent_BlackorHispanic = mean (pctBlackorHis, na.rom=T),
            Percent_Bach = mean(pctBachelors, na.rm = T),
            Percent_Poverty = mean(pctPoverty, na.rm = T),
            Percent_CarCommute = mean(pctCarCommute, na.rm = T),
            Percent_PubCommute = mean(pctPubCommute, na.rm = T),
            Median_Income = mean(MedInc, na.rm = T))
            
kable(allTracts.Summary) %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Summary for All Tracts in Washington, DC")

allTracts.Summary %>%
  unite(year.TOD, year, TOD, sep = ": ", remove = T) %>%
  gather(Variable, Value, -year.TOD) %>%
  mutate(Value = round(Value, 2)) %>%
  spread(year.TOD, Value) %>%
  kable() %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Summary for All Tracts in Washington, DC")
allTracts.Summary
```

# Grouped Bar Graphs: Comparing Selected Census Variables across Time and Space (TOD vs. non-TOD)
```{r comparison_groupedbars}
allTracts.Summary %>%
  gather(Variable, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Variable, scales = "free", ncol=3) +
  scale_fill_manual(values = paletteTOD) +
  labs(title = "Indicator differences across time and space") +
  plotTheme() + theme(legend.position="bottom")

```

# Graduated Symbol Maps
intro text
```{r centroid_wrangling_TOD_filter}

# More wrangling to get centroids for all tracts, both years, filtering by TOD
allTracts.group.TODonly <-
  filter(allTracts.group, TOD =="TOD")

allTracts.group.TODonly.centroids <- sf::st_centroid(allTracts.group.TODonly) %>%
  dplyr::mutate(lat = sf::st_coordinates(.)[,1],
                lon = sf::st_coordinates(.)[,2])
allTracts.group.TODonly.centroids
```

## Graduated symbol map of population within 1/2 mi of each wmata station
```{r}

```

## Graduated symbol map of median rent within 1/2 mi of each wmata station
```{r}

```

# Crime section
Intro text
## Crime data wrangling
```{r}
#2009 Crime data wrangling
```
```{r}
#2017 Crime data wrangling
```
```{r}
#Some code and some analyses, etc
```
## Multiple ring buffer


